Manuscript

### 2022-06-30 Manuscript.Rmd
updated rmd yaml for word doc export. added initial .bib file for references.

---
title: "Manuscript"
author: "Justin Pomeranz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \newcommand{\beginsupplement}{ \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
output:
  pdf_document: 
    toc: yes
  html_document: default
---

@@

@@

Body size distributions are a fundamental characteristic of communities. The remarkable consistency of these relationships across spatiotemporal scales and habitat types has led them to be recommended as a "universal" indicator of ecological status (Petchey and Belgrano 2010). Size metrics have commonly been used in marine systems, and are increasingly being applied to assess the condition of freshwater ecosystems (Martinez et al. 2016, Pomeranz 2018).

@@

Previous work has shown that the estimates of $\lambda$ differ between MLE and OLS techniques. OLS methods are particularly sensitive to decisions made in the binning process. Simulation studies have shown that MLE offers consistently more accurate estimates of $\lambda$ (White et al. 2007, Edwards et al. 2017), and reanalysis of empirical datasets also indicates that the conclusions are dependent on the methodology used (White et al. 2007, Edwards et al. 2020). However, recent empirical analysis of stream macroinvertebrate communities across the NAtional Ecological Observatory Network (NEON, USA) showed that while the estimates of $\lambda$ varied, the relative change across the environmental gradient was consistent regardless of method used (Pomeranz et al. 2022). While there is a growing consensus that MLE methods offer more reliable estimates of $\lambda$, and binning methods result in biased estimates, it remains unclear if these biases are consistent and systematic or stochastic, and whether or not the relative change in ISD parameters is consistent across space and time. In other words, if the data within a study are all treated the same, does a relative change of OLS slope parameters of 0.1 coincide with a relative change of MLE estimates of 0.1? In order to answer this question, we simulate body size observations from bounded power law distributions with varied $\lambda$ exponents across a hypothetical environmental gradient and compare the results obtained using three different methodologies common in the literature. In addition, we re-analyze two previously published datasets of stream community body sizes across a stress and environmental gradient. We find that the MLE method more accurately estimates the site-specific $\lambda$ exponents as well as the relative change in exponents across the hypothetical gradient, as well as having smaller variation. The logarthmic binning methods generally perform well, but have larger confidence intervals around the estimates. However, the overall conclusions of previously published empirical data sets are not dependent on the method used.

@@

In order to investigate the performance of commonly used methods, we simulate body size observations from a bounded power law distribution using the inverse method, as described in Edwards et al. (2017). Let $M$ be a random variable of body sizes described by the probability density function:

@@

$M_{min} = 0.0026$ and $M_{max} = 1.2 *10^3$. These values are based on empirical body sizes of stream benthic communities reported in Pomeranz et al. (2020). Our results are not dependent on the range of body sizes, and we show the results of other ranges in the supplemental information.

@@

For each sample of $M$ and each of the 1000 replicates, we estimated the exponent $\lambda$ using MLE methods modified from the the `sizeSpectra` package (Edwards et al. 2020). In addition, we use two common binning methods to estimate the OLS slope parameter in log-log space. For the first binning method, we created 6 equal logarithmic bins covering the range of body sizes. The count in each bin was normalized by dividing by the bin width. This method has been used by (refs here, I think a lot of Woodward's group uses this...) Throughout the manuscript, the normalized equal logarithmic binning method will be referred to as ELBn. The second method was similar to ELBn, but bins of Log~2~ widths are used. The count in each bin is normalized in the same way. This method has been used by (Pomeranz et al. 2018, I **think** McGarvey \~2018, Blanchard (log 2, ?normalized?)) and is referred to as the Normalized Abundance Spectrum (NAS).

@@

We re-analyze two datasets of benthic macroinvertebrate communities from stream habitats across two different gradients. In the first, quantitative macroinvertebrate samples were collected from streams across an acid mine drainage stress gradient. Details of the sample collection and processing can be found in Pomeranz et al. (2018). Briefly, all individuals from each sample were identified to the lowest practical taxonomic unit and body lengths were measured using image processing software from photos taken with a camera mounted to a dissecting microscope. Body mass was estimated using published length weight regressions.

The second data set was from the wadeable stream sites of the National Ecological Observatory Network (NEON; **data product XXX**). These sites are located across a wide temperature gradient in the United States, from Puerto Rico to Alaska. Quantitative macroinvertebrate samples were collected using the most appropriate method based on the local habitat. All individuals were identified and had their body lengths measured, and body mass was estimated using published length weight regressions. This data has been analyzed previously using size spectra methods as described in Pomeranz et al. (2022). Detailed methods of the sampling collection and data processing methods can be found on the NEON website (URL, or macroinvertebrate DPI pubs).

@@

*alternative* absolute difference of \~0.1 units in $\lambda$ across gradients is similar to differences reported in other studies (seasonally: 0.16, Mcgarvey and Kirk; landuse: 0.15 Martinez; O'Gorman 2017 \~0.25, Yvon and Dossena papers \~0.1? [need to redownload some of these and compare])

@@

We reiterate the recommendations of White et al. (2007), Sprules and Barth (2016) and Edwards et al. (2017) to estimate size spectra relationships using MLE methods due to their superior performance in nearly every context. Furthermore, we strongly encourage authors to publish individual size data whenever possible. This will allow for the consistent re-analysis of existing data sets as methodologies develop and improve. This will aid in the ability for size spectra work to be synthesized between research groups and across scales.

>>>>

---
title: "Manuscript"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: 
  - Justin Pomeranz:
      email: jfpomeranz@gmail.com
      institute: cmu
      correspondence: true
  - name: James R. Junker
    institute: [glrc, lumcon]
  - name: Jeff S. Wesner
    institute: usd
institute:
  - cmu: Colorado Mesa University
  - glrc: Great Lakes Research Center, Michigan Technological University, Houghton, MI USA
  - lumcon: Louisiana Universities Marine Consortium, Chauvin, LA USA
  - usd: Dept. of Biology, University of South Dakota, Vermillion, SD, USA
output:
   pdf_document:
     toc: yes
     keep_tex: true
     pandoc_args:
     - '--lua-filter=scholarly-metadata.lua'
     - '--lua-filter=author-info-blocks.lua'
   word_document:
      reference_docx: working_docx_template.docx
      toc: no
      pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
   html_document: default
      pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
header-includes:
  - \usepackage{lineno}
  - \usepackage{amsmath}
  - \usepackage{indentfirst}
  - \linenumbers
  - \newcommand{\beginsupplement}{ \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
indent: true
linestretch: 1
bibliography: refs.bib
link-citations: no
link-color: grey
csl: ecology.csl
---

@@

**NOTE from Jim** I use Zotero, mainly for free and open-source reasons, but I also can have a standalone program that doesn't require internet access, can backup my reference library on the cloud and sync among multiple computers, can spit out bibtex to connect to markdown, and can have group citation lists if neccessary. I use it with a betterBibtex add on that allows for modifying bibtex keys for calling refs in Rmarkdown.

@@

Body size distributions are a fundamental characteristic of communities. The remarkable consistency of these relationships across spatiotemporal scales and habitat types has led them to be recommended as a "universal" indicator of ecological status (Petchey and Belgrano 2010)[@petchey2010]. Size metrics have commonly been used in marine systems, and are increasingly being applied to assess the condition of freshwater ecosystems (Martinez et al. 2016, Pomeranz 2018)[@martinez2016; @pomeranz2018].

@@

Previous work has shown that the estimates of $\lambda$ differ between MLE and OLS techniques. OLS methods are particularly sensitive to decisions made in the binning process. Simulation studies have shown that MLE offers consistently more accurate estimates of $\lambda$ (White et al. 2007, Edwards et al. 2017)[@white2007; @edwards2017], and reanalysis of empirical datasets also indicates that the conclusions are dependent on the methodology used (White et al. 2007, Edwards et al. 2020)[@white2007; @edwards2020]. However, recent empirical analysis of stream macroinvertebrate communities across the NAtional Ecological Observatory Network (NEON, USA) showed that while the estimates of $\lambda$ varied, the relative change across the environmental gradient was consistent regardless of method used (Pomeranz et al. 2022)[@pomeranz2022]. While there is a growing consensus that MLE methods offer more reliable estimates of $\lambda$, and binning methods result in biased estimates, it remains unclear if these biases are consistent and systematic or stochastic, and whether or not the relative change in ISD parameters is consistent across space and time. In other words, if the data within a study are all treated the same, does a relative change of OLS slope parameters of 0.1 coincide with a relative change of MLE estimates of 0.1? In order to answer this question, we simulate body size observations from bounded power law distributions with varied $\lambda$ exponents across a hypothetical environmental gradient and compare the results obtained using three different methodologies common in the literature. In addition, we re-analyze two previously published datasets of stream community body sizes across a stress and environmental gradient. We find that the MLE method more accurately estimates the site-specific $\lambda$ exponents as well as the relative change in exponents across the hypothetical gradient, as well as having smaller variation. The logarthmic binning methods generally perform well, but have larger confidence intervals around the estimates. However, the overall conclusions of previously published empirical data sets are not dependent on the method used.

@@

In order to investigate the performance of commonly used methods, we simulate body size observations from a bounded power law distribution using the inverse method, as described in Edwards et al. (2017)[-@edwards2017]. Let $M$ be a random variable of body sizes described by the probability density function:

@@

$M_{min} = 0.0026$ and $M_{max} = 1.2 *10^3$. These values are based on empirical body sizes of stream benthic communities reported in Pomeranz et al. (2020)[@pomeranz2020]. Our results are not dependent on the range of body sizes, and we show the results of other ranges in the supplemental information.

@@

For each sample of $M$ and each of the 1000 replicates, we estimated the exponent $\lambda$ using MLE methods modified from the the `sizeSpectra` package (Edwards et al. 2020). In addition, we use two common binning methods to estimate the OLS slope parameter in log-log space. For the first binning method, we created 6 equal logarithmic bins covering the range of body sizes. The count in each bin was normalized by dividing by the bin width. This method has been used by (refs here, I think a lot of Woodward's group uses this...) Throughout the manuscript, the normalized equal logarithmic binning method will be referred to as ELBn. The second method was similar to ELBn, but bins of Log~2~ widths are used. The count in each bin is normalized in the same way. This method has been used by (Pomeranz et al. 2018, I **think** McGarvey \~2018, Blanchard (log 2, ?normalized?))[@pomeranz2018; @mcgarvey2019] and is referred to as the Normalized Abundance Spectrum (NAS).

@@

We re-analyze two datasets of benthic macroinvertebrate communities from stream habitats across two different gradients. In the first, quantitative macroinvertebrate samples were collected from streams across an acid mine drainage stress gradient. Details of the sample collection and processing can be found in Pomeranz et al. (2018)[@pomeranz2018]. Briefly, all individuals from each sample were identified to the lowest practical taxonomic unit and body lengths were measured using image processing software from photos taken with a camera mounted to a dissecting microscope. Body mass was estimated using published length weight regressions.

The second data set was from the wadeable stream sites of the National Ecological Observatory Network (NEON; **data product XXX**)[@NEON_Inverts2022]. These sites are located across a wide temperature gradient in the United States, from Puerto Rico to Alaska. Quantitative macroinvertebrate samples were collected using the most appropriate method based on the local habitat. All individuals were identified and had their body lengths measured, and body mass was estimated using published length weight regressions. This data has been analyzed previously using size spectra methods as described in Pomeranz et al. (2022)[@pomeranz2022]. Detailed methods of the sampling collection and data processing methods can be found on the NEON website ([https://www.neonscience.org/data-collection/macroinvertebrates], or macroinvertebrate DPI pubs).

@@

*alternative* absolute difference of \~0.1 units in $\lambda$ across gradients is similar to differences reported in other studies (seasonally: 0.16, Mcgarvey and Kirk; landuse: 0.15 Martinez; O'Gorman 2017 \~0.25, Yvon and Dossena papers \~0.1? [need to redownload some of these and compare])[@mcgarvey2018; @dossena2012; @ogorman2017]

@@

We reiterate the recommendations of White et al. (2007)[-@white2007], Sprules and Barth (2016)[-@sprules2016] and Edwards et al. (2017)[-@edwards2017] to estimate size spectra relationships using MLE methods due to their superior performance in nearly every context. Furthermore, we strongly encourage authors to publish individual size data whenever possible. This will allow for the consistent re-analysis of existing data sets as methodologies develop and improve. This will aid in the ability for size spectra work to be synthesized between research groups and across scales.

# References

<div id="refs"></div>
